---
title: "dust: Random number generation"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{dust: Random number generation}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  error = FALSE,
  collapse = TRUE,
  comment = "#>",
  fig.width = 7,
  fig.height = 5)
```

We provide an interface to the "Xoshiro256" generators (Xoshiro is derived from XOR/shift/rotate). These are designed to allow use in parallel by "jumping ahead" in the sequence and we use this below to interleave generators. The stream is unrelated to and unaffected by R's random number generation. `set.seed` has no effect, for example. **The random numbers are not cryptographically secure**; for that see the excellent [sodium](https://cran.r-project.org/package=sodium) package.

Ordinarily this is used from C++; the model as discussed in `vignette("dust")` uses `dust::rng_state_t<real_t>` and a function from `dust::rand` to interact with the generator. However, an R interface is provided for debugging and testing purposes.

```{r}
rng <- dust::dust_rng$new(seed = 1)
rng
```

Currently only a few distributions are supported, with an interface that mimics R's interface both in names of functions and of arguments, with the exception that there are no default arguments:

```{r}
rng$rnorm(100, 0, 1)
```

One feature that we use here is to allow multiple streams of random numbers within a rng object. If running in parallel (via a dust model) these different random number streams can be given to different threads.

```{r}
rng1 <- dust::dust_rng$new(seed = 1, n_generators = 1)
rng2 <- dust::dust_rng$new(seed = 1, n_generators = 2)
rng1$unif_rand(5)
rng2$unif_rand(10)
```

Notice here how in the output from `rng2`, values 1, 3, ..., 9 correspond to the 5 numbers out of `rng1`.

This is achieved by "jumping" the random number streams forward.  Here are the random numbers 2, 4, ..., 10 from the output of rng2:

```{r}
rng3 <- dust::dust_rng$new(seed = 1, n_generators = 1)$jump()
rng3$unif_rand(5)
```

A jump is equivalent to 2^128 draws from the random number generator (about 10^38). There are 2^128 of these non-overlapping subsequences in the generator, which is quite a lot. If this feels too close together, then the `$long_jump()` method jumps even further (2^192 draws, or about 10^57). There are 2^64 (10^20) of these sequences.

## Supported distributions

We do not yet support the full set of distributions provided by R. Like with R, the *R* interface to these functions takes an argument `n` for the number of samples but the C++ interface omits this argument

**Uniform distribution** between `min` and `max`:

```{r}
rng$runif(10, 3, 6)
```

From C++

```cc
real_t u = rng.unif(3, 6);
```

**Normal distribution** with parameters `mean` and `sd`

```{r}
rng$rnorm(10, 3, 6)
```

```cc
real_t z = rng.rnorm(3, 6);
```

**Poisson distribution** with mean `lambda`

```{r}
rng$rpois(10, 4.5)
```

```cc
int n = rng.rpois(4.5);
```

**Binomial distribution** with mean `size` and `prob`

```{r}
rng$rbinom(10L, 10L, 0.3)
```

```cc
int n = rng.rbinom(10, 0.3);
```

There are also helpers for some common special case distributions:

**Standard uniform distribution** (between 0 and 1) - faster than using `$runif(n, 0, 1)`:

```{r}
rng$unif_rand(10)
```

```cc
real_t u = rng.unif_rand();
```

**Standard normal distribution** (mean 0 and standard deviation 1) - faster than using `$rnorm(n, 0, 1)`:

```{r}
rng$norm_rand(10)
```

```cc
real_t z = rng.norm_rand();
```

Supported distributions are documented more-fully in `?dust::dust_rng`.

## Performance

Performance should be on par with R's random number generator, though here the timings are likely to be mostly due to allocations and copies of memory:

```{r}
bench::mark(
  rng1$unif_rand(1000),
  rng1$runif(1000, 0, 1),
  runif(1000),
  time_unit = "us",
  check = FALSE)
```

The difference between `unif_rand` and `runif` here is the cost of recycling the parameters, not the actual generation!

Binomial distribution, small `n * p`, which uses an inversion algorithm

```{r}
rng1 <- dust::dust_rng$new(seed = 1, n_generators = 1)
n <- rep(9:10, length.out = 1000)
p <- rep(c(0.1, 0.11), length.out = 1000)
bench::mark(
  rng1$rbinom(1000, n, p),
  rbinom(1000, n, p),
  time_unit = "us",
  check = FALSE)
```

(note we vary `n` and `p` here as we've optimised things for random parameter access).

```{r}
n <- rep(9999:10000, length.out = 1000)
p <- rep(c(0.3, 0.31), length.out = 1000)
bench::mark(
  rng1$rbinom(1000, n, p),
  rbinom(1000, n, p),
  time_unit = "us",
  check = FALSE)
```

Practically, performance will vary based on the parameters of the distributions and the underlying algorithms, and the principle performance gain we hope to get here is driven by the ability to parallelise safely rather than the speed of the draws themselves.
